alluvial diagrams in ggplot2

### 2017-10-22 inst/doc/ggalluvial.rmd
edit vignette

- An *axis* is a dimension (variable) along which the data are vertically grouped at a fixed horizontal position. The diagram above uses three categorical axes: `Class`, `Sex`, and `Age`.
- The groups at each axis are depicted as opaque blocks called *strata*. For example, the `Class` axis contains four strata: `1st`, `2nd`, `3rd`, and `Crew`.
- Horizontal (x-) splines called *alluvia* span the width of the diagram. In this diagram, each alluvium corresponds to a fixed value of each axis variable, indicated by its vertical position at the axis, as well as of the `Survived` variable, indicated by its fill color.
- The segments of the alluvia between pairs of adjacent axes are *flows*.
- The alluvia intersect the strata at *lodes*, The lodes are not visualized in the above diagram, but they can be inferred as filled rectangles extending the flows through the strata at each end of the diagram or connecting the flows on either side of the center stratum.

@@

**ggalluvial** recognizes two formats of "alluvial data", detailed in the following subsections. A third, tabular, form is popular for storing categorical data such as the `Titanic` and `UCBAdmissions` datasets used in this vignette.[^tableform] For consistency with tidy data principles and **ggplot2** conventions, **ggalluvial** does not accept tabular input.

@@

This form was inherited from the first version of **ggalluvial**, which modeled it after usage in **alluvial**. It required a stark departure from the usual position aesthetics: The user declares any number of axis variables, which `stat_alluvium()` and `stat_stratum()` recognize and process in a consistent way:

@@

An important feature of these diagrams is the meaningfulness of the vertical axis: No gaps are inserted between the strata, so the total height of the diagram reflects the cumulative weight of the observations. The diagrams produced by **ggalluvial** conform (somewhat; see below) to the "grammar of graphics" principles of **ggplot2**, and this prevents users from producing "free-floating" diagrams like the Sankey diagrams showcased [here](https://developers.google.com/chart/interactive/docs/gallery/sankey).
**ggalluvial** parameters and existing **ggplot2** functionality can also produce [parallel sets](https://eagereyes.org/parallel-sets) plots, illustrated here using the `Titanic` dataset:[^ggparallel]

@@

- The `axis[0-9]*` position aesthetics are non-standard.
- `stat_alluvium()` ignores any argument to the `group` aesthetic; instead, `StatAlluvium$compute_panel()` uses `group` to link the rows of the internally-transformed dataset that correspond to the same alluvium.
- The `label.strata` parameter instructs `stat_stratum()` (called by `geom_text()`) to take the values of the axis variables as labels.
- The horizontal axis must be manually corrected (using `scale_x_continuous()`) to reflect the implicit categorical variable identifying the axis.

@@

The second format recognized by **ggalluvial** contains one row per lode, and can be understood as the result of "gathering" (in the **dplyr** sense) the axis columns of a dataset of the preceding kind into a "key" column indicating the axis and a "value" column indicating the value taken thereat. This format requires an additional indexing column that links the rows corresponding to a common alluvium, as illustrated below using the `to_lodes()` defaults on the Berkeley admissions dataset:

@@

- `x`, the "key" variable indicating the axis to which the row corresponds, which are to be arranged along the horizontal axis;
- `stratum`, the "value" taken by the axis variable indicated by `x`; and
- `alluvium`, the indexing scheme that links the rows of a single alluvium.

Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and usually not plotted. For convenience, both `stat_alluvium()` and `stat_flow()` accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can grop countries in the `Refugees` dataset by region, in order to compare refugee volumes hierarchically:

@@

The format allows us to assign aesthetics that change from axis to axis along the same alluvium, which is useful for repeated measures datasets. This requires generating a separate graphical object for each flow, as implemented in `geom_flow()`.
The plot below uses a set of (changes to) students' academic curricula over the course of several semesters.
Since `geom_flow()` calls `stat_flow()` by default (see the next example), we override it with `stat_alluvium()` in order to track each student across all semesters:

@@

No `weight` is specified, so each row is assigned a weight of `1`.
This example demonstrates one way **ggalluvial** handles missing data. The alternative is to set the parameter `na.rm` to `TRUE`.[^na.rm] Missing data handling (specifically, the order of the strata) also depends on whether the `stratum` variable is character or factor/numeric.

@@

This diagram ignores any continuity between the alluvial segments linking different pairs of adjacent axes. This "memoryless" plot produces a less cluttered diagram, in which at most one flow proceeds from each stratum at one axis to each stratum at the next, but at the cost of being able to track each cohort across the entire diagram.

>>>>

- An *axis* is a dimension (variable) along which the data are vertically grouped at a fixed horizontal position. The diagram above uses three categorical axes: `Class`, `Sex`, and `Age`.
- The groups at each axis are depicted as opaque blocks called *strata*. For example, the `Class` axis contains four strata: `1st`, `2nd`, `3rd`, and `Crew`.
- Horizontal (x-) splines called *alluvia* span the width of the diagram. In this diagram, each alluvium corresponds to a fixed value of each axis variable, indicated by its vertical position at the axis, as well as of the `Survived` variable, indicated by its fill color.
- The segments of the alluvia between pairs of adjacent axes are *flows*.
- The alluvia intersect the strata at *lodes*, rectangles extending the flows through the strata at each end of the diagram or connecting the flows on either side of the center stratum. The lodes are generated as parts of the alluvia in the above diagram, but they are obscured from view by the strata.

@@

**ggalluvial** recognizes two formats of "alluvial data", in which rows record either alluvia or lodes. The formats are related by a pair of operations, `to_lodes()` and `to_alluvia()`. These functions are adapted from `tidyr::spread()` and `tidyr::gather()` in order to also maintain the entanglement between alluvia and strata.
A third, tabular, form is popular for storing categorical data such as the `Titanic` and `UCBAdmissions` datasets used in this vignette.[^tableform] For consistency with tidy data principles and **ggplot2** conventions, **ggalluvial** does not accept tabular input.

@@

This form was inherited from the first version of **ggalluvial**, which modeled it after usage in **alluvial**. It required a stark departure from the usual position aesthetics: The user declares any number of axis variables, which `stat_alluvium()` and `stat_stratum()` must recognize and process in a consistent way:

@@

The vertical axis scale is an important feature of these diagrams: No gaps are inserted between the strata, so the total height of the diagram reflects the cumulative weight (in this case, the count) of the observations. The diagrams produced by **ggalluvial** conform (somewhat; see below) to the "grammar of graphics" principles of **ggplot2**, and this prevents users from producing "free-floating" diagrams like the Sankey diagrams showcased [here](https://developers.google.com/chart/interactive/docs/gallery/sankey).

**ggalluvial** parameters and existing **ggplot2** functionality can also produce [parallel sets](https://eagereyes.org/parallel-sets) plots, illustrated here using the `Titanic` dataset:[^ggparallel]

@@

- The `axis[0-9]*` position aesthetics are non-standard.
- `stat_alluvium()` ignores any argument to the `group` aesthetic; instead, `StatAlluvium$compute_panel()` uses `group` to link the rows of the internally-transformed dataset that correspond to the same alluvium.
- The `label.strata` parameter instructs `stat_stratum()` (called by `geom_text()` in the above code chunk) to take the values of the axis variables as labels.
- The horizontal scale must be manually corrected (using `scale_x_continuous()`) to reflect the implicit categorical variable indicating which axis variable each horizontal position corresponds to.

@@

The second format recognized by **ggalluvial** contains one row per lode, and can be understood as the result of `tidyr::gather()`ing the axis columns of a dataset of the preceding kind into a "key" column indicating the axis and a "value" column indicating the value taken thereat. This format requires an additional indexing column that links the rows corresponding to a common alluvium, as illustrated below using the `to_lodes()` defaults on the Berkeley admissions dataset:

@@

- `x`, the "key" variable indicating the axis to which the row corresponds, which are to be arranged horizontally;
- `stratum`, the "value" taken by the axis variable indicated by `x`; and
- `alluvium`, the indexing scheme that links the lodes of a single alluvium.

Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and usually not plotted. For convenience, both `stat_alluvium()` and `stat_flow()` accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can group countries in the `Refugees` dataset by region, in order to compare refugee volumes hierarchically:

@@

The format allows us to assign aesthetics that change from axis to axis along an alluvium, which is useful for repeated measures datasets. This requires generating a separate graphical object for each flow, as implemented in `geom_flow()`.
The plot below visualizes a set of (changes to) students' academic curricula over the course of several semesters.
Since `geom_flow()` calls `stat_flow()` by default (see the next example), we override it with `stat_alluvium()` in order to track each student across all semesters:

@@

No `weight` is specified, so each row is assigned unit weight.
This example demonstrates one way **ggalluvial** handles missing data. The alternative is to set the parameter `na.rm` to `TRUE`.[^na.rm] Missing data handling (specifically, the order of the strata) also depends on whether the `stratum` variable is character or factor/numeric.

@@

This diagram ignores any continuity between the alluvial segments linking different pairs of adjacent axes. This "memoryless" plot produces a less cluttered diagram, in which at most one flow proceeds from each stratum at one axis to each stratum at the next, but at the cost of being able to track each individual or group longitudinally across the diagram.
